---
title: "Introduction to socialmixr"
author: "Sebastian Funk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to socialmixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library('knitr')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[socialmixr](https://github.com/sbfnk/socialmixr) is an `R` package to derive social mixing matrices from survey data. These are particularly useful for age-structured [infectious disease models](https://en.wikipedia.org/wiki/Mathematical_modelling_of_infectious_disease). For background on age-specific mixing matrices and what data inform them, see, for example, the paper on POLYMOD by [Mossong et al](https://doi.org/10.1371/journal.pmed.0050074).

Usage
======

The latest stable version of the `socialmixr` package is installed via
```{r eval=FALSE}
install.packages('socialmixr')
```

The latest development version of the `socialmixr` package can be installed via
```{r eval=FALSE}
devtools::install_github('sbfnk/socialmixr')
```

To load the package, use
```{r eval=FALSE}
library('socialmixr')
```
```{r echo=FALSE}
suppressWarnings(library('socialmixr'))
```


At the heart of the `socialmixr` package is the `contact_matrix` function. This extracts a contact matrix from survey data. You can use the `R` help to find out about usage of the `contact_matrix` function, including a list of examples:

```{r eval=FALSE}
?contact_matrix
```

The POLYMOD data are included with the package and can be loaded using

```{r}
data(polymod)
```

An example use would be

```{r}
contact_matrix(polymod, countries = "United Kingdom", age.limits = c(0, 1, 5, 15))
```

This generates a contact matrix from the UK part of the POLYMOD study, with age groups 0-1, 1-5, 5-15 and 15+ years. It contains the mean number of contacts that each member of an age group (row) has reported with members of the same or another age group (column).

Methodology
==============

The `contact_matrix` function requires a survey given as a list of two elements, both given as data.frames: `participants` and `contacts`. They must be linked by an ID column that refers to the identity of the queried participants (by default `global_id`, but this can be changed using the `id.column` argument). The `participants` data frame, as a minimum, must have the ID column and a column denoting participant age (which can be set by the `part.age.column` argument, by default `participant_age`). The `contacts` data frame, similarly, must have the ID column and a column denoting age (which can be set by the `contact.age.column` argument, by default `cnt_age_mean`).

The function then either randomly samples participants (if `bootstrap` is set to `TRUE`) or takes all participants in the survey and determines the mean number of contacts in each age group given the age group of the participant. The age groups can be set using the `age.limits` argument, which should be set to the lower limits of the age groups (e.g., `age.limits=c(0, 5, 10)` for age groups 0-5, 5-10, 10+). If these are not given, the narrowest age groups possible given survey and demographic data are used.

Surveys
=======

The key argument to the `contact_matrix` function is the `survey` that it supposed to use. The `socialmixr` package includes the POLYMOD survey, which will be used if not survey is specified. It also provides access to all surveys in the [Social contact data](https://zenodo.org/communities/social_contact_data) community on [Zenodo](https://zenodo.org). The available surveys can be listed (if an internet connection is available) with

```{r eval=FALSE}
list_surveys()
```

A survey can be downloaded using the `get_survey` command. This will get the relevant data of a survey given its Zenodo DOI (as returned by `list_surveys`). All other relevant commands in the `socialmixr` package accept a DOI, but if a survey is to be used repeatedly it is worth downloading it and storing it locally to avoid the need for a network connection and speed up processing.

```{r eval=FALSE}
peru_survey <- get_survey("https://doi.org/10.5281/zenodo.1095664")
saveRDS(peru_survey, "peru.rds")
```

This way, the `peru` data set can be loaded in the future without the need for an internet connection using
```{r eval=FALSE}
peru_survey <- readRDS("peru.rds")
```

Some surveys may contain data from multiple countries. To check this, use the `survey_countries` function

```{r}
survey_countries(polymod)
```

If one wishes to get a contact matrix for one or more specific countries, a `countries` argument can be passed to `contact_matrix`. If this is not done, the different surveys contained in a dataset are combined as if they were one single sample (i.e., not applying any population-weighting by country or other correction).

By default, socialmixr uses the POLYMOD survey. A reference for any given survey can be obtained using `cite`, e.g.

```{r}
cite(polymod)
```

Bootstrapping
=======

To get an idea of uncertainty of the contact matrices, a bootstrap can be used. If an argument `n` greater than 1 is passed to `contact_matrix`, multiple samples of contact matrices are generated. For each sample, participants are sampled (with replacement, to get the same number of participants of the original study), and contacts are sampled from the set of all the contacts of all the participants (again, with replacement). All resulting contact matrices are returned as `matrices` field in the returned list. From these, derived quantities can be obtained, for example the mean

```{r}
m <- contact_matrix(polymod, countries = "United Kingdom", age.limits = c(0, 1, 5, 15), n=5)
length(m$matrices)
mr <- Reduce("+", lapply(m$matrices, function(x) {x$matrix})) / length(m$matrices)
mr
```

Demography
=======

Obtaining symmetric contact matrices or splitting out their components (see below) requires information about the underlying demographic composition of the survey population. This can be passed to `contact_matrix` as the `survey.pop` argument, a `data.frame` with two columns, `lower.age.limit` (denoting the lower end of the age groups) and `population` (denoting the number of people in each age group). If no `survey.pop` is not given, `contact_matrix` will try to obtain the age structure of the population (as per the `countries` argument) from the [World Population Prospects](https://esa.un.org/unpd/wpp/) of the United Nations, using estimates from the year that closest matches the year in which the contact survey was conducted.

If demographic information is used, this is returned by `contact_matrix` as the `demography` field in the results list. It is possible to enforce or prevent the function to return demography data by using the boolean `return.demography`.

Symmetric contact matrices
=======

Conceivably, contact matrices should be symmetric: the total number of contacts made by members of one age group with those of another should be the same as vice versa. Mathematically, if $c_{ij}$ is the mean number of contacts made by members of age group $i$ with members of age group $j$, and the total number of people in age group $i$ is $N_i$, then

$$c_{ij} N_i = c_{ji}N_j$$

Because of variation in the sample from which the contact matrix is obtained, this relationship is usually not fulfilled exactly. In order to obtain a symmetric contact matrix that fulfills it, one can use

$$c'_{ij} = \frac{1}{2N_i} (c_{ij} N_i + c_{ji} N_j)$$

To get this version of the contact matrix, use `symmetric = TRUE` when calling the `contact_matrix` function.

```{r}
contact_matrix(polymod, countries = "United Kingdom", age.limits = c(0, 1, 5, 15), symmetric = TRUE)
```

Splitting contact matrices
=======

The `contact_matrix` contains a simple model for the elements of the contact matrix, by which it is split into a _global_ component, as well as three components representing _contacts_, _assortativity_ and _demography_. In other words, the elements $c_{ij}$ of the contact matrix are modelled as

$$ c_{ij} = q d_i a_{ij} n_j $$

where $q d_i$ is the number of contacts that a member of group $i$ makes across age groups, $n_j$ is the proportion of the surveyed population in age group $j$. The constant $q$ is set to the value of the largest eigenvalue of $c_{ij}$; if used in an infectious disease model, it can be replaced by the basic reproduction number $R_0$.

To model the contact matrix in this way with the `contact_matrix` function, set `split = TRUE`. The components of the matrix are returned as elements `normalisation` ($q$), `contacts` ($d_i$), `matrix` ($a_{ij}$) and `demography` ($n_j$) of the resulting list.

```{r}
contact_matrix(polymod, countries = "United Kingdom", age.limits = c(0, 1, 5, 15), split = TRUE)
```

Plotting
=======

The contact matrices can be plotted, for example, using the `geom_tile` function of the `ggplot2` package.

```{r fig.width=5, fig.height=5, tidy=TRUE}
library('reshape2')
library('ggplot2')
df <- melt(mr, varnames = c("age1", "age2"), value.name = "contacts")
ggplot(df, aes(x = age2, y = age1, fill = contacts)) + theme(legend.position="bottom") + geom_tile()
```

Participant weights
=======


Participant weights are commonly used to align sample and population characteristics in terms of temporal aspects and the age distribution. For example, the day of the week has been reported as a driving factor for social contact behavior, hence to obtain a weekly average, the survey data should represent the weekly 2/5 distribution of weekend/week days. To align the survey data to this distribution, one can obtain participant weights in the form of:
$$w_{day.of.week} = \tfrac{5/7}{N_{weekday}/N} \text{  OR   } \tfrac{2/7}{N_{weekend}/N}$$
with sample size $N$, and $N_{weekday}$ and $N_{weekend}$ the number of participants that were surveyed during weekdays and weekend days, respectively. It is possible to remove the constant values (e.g. $w = 5/N_{weekday}$), which results in the same standardized weights. However, we opt to use the relative proportions to calculate weights to enable truncation with a generic threshold (see below). 

Another driver of social contact patterns is age. To improve the representativeness of survey data, age-specific weights can be calculated as:
$$w_{age} = \tfrac{P_{a}/P}{N_{a}/N}$$
with $P$ the population size, $P_a$ the population fraction of age $a$, $N$ the survey sample size and $N_a$ the survey fraction of age $a$. The combination of age-specific and temporal weights for participant $i$ of age $a$ can be constructed as:
$$w_{i} = w_{age} * w_{day.of.week} $$
Finally, the weights can to be standardised as follows:
$$\tilde{w}_{i} =  \tfrac{w_{i}}{\sum_{}^{} w_{}} * N $$
However, if the social contact analyses is based on (age) stratification by splitting the population into non-overlapping groups (strata), it requires the weights to be standardized so that the weighted totals within mutually exclusive cells equal the known population totals (Kolenikov 2016). The post-stratification cells need to be mutually exclusive and cover the whole population. The post-stratified (PS) weight for participant $i$ of \mbox{group $g$} is:
$$\tilde{w}^{PS}_{i} =  \tfrac{w_{i}}{\sum_{\text{j}}^{\text{group g}} w_{j}} * N_g$$

Temporal weights are activated in `contact_matirx` by `weigh.dayofweek` and age-specific weights are used when enabling `weight.age`. The post-stratification weights are calculated by default. It is possible to obtain the participant weights via the option `return.part.weights`.

```{r message=FALSE, warning=FALSE}
contact_matrix(survey = polymod, age.limits = c(0,18,60), weigh.dayofweek = TRUE, 
               weigh.age = TRUE, return.part.weights = TRUE)

```


Participant weights threshold
=======

If the survey population differs extensively from the demography, (some) participants can end up with high weights and as such, a large contribution to the population averages. It can be warranted to limit the influence of single participants by a truncation of the weights. This truncation is applied on the standardized weights,  followed by another standardisation to make sure that the sum of the weights still equals the sample size. The latter can lead to final weights of which some little exceed the threshold value. To enable the threshold in `contact_matrix`, you need to provide a numeric `weight.threshold`.

```{r message=FALSE, warning=FALSE}
contact_matrix(survey = polymod, age.limits = c(0,18,60), weigh.dayofweek = TRUE, 
               weigh.age = TRUE, return.part.weights = TRUE,
               weight.threshold = 3)
```

Contact rates per capita
=======

Contact matrices $c_{ij}$ represent typically the reported number of contacts of an individual of age $i$ with individuals of age $j$. The contact matrix per capita provides the contact rates with one individual age $j$ in the reference population demographics. These per capita matrices can be used in infectious disease modelling to calculate for example the force of infection. The latter is based on the likelihood that one susceptible individual of age $i$ will be in contact with one infectious individual of age $j$. 

The contact rates per capita are calculated as follows:

$$\tilde{c}_{ij} =  \tfrac{c_{ij}}{N_{j}}$$

Please note that if the option `symmetric=TRUE` is specified in `contact_matrix`, the matrices per capita will actually be symmetric, while the default contact matrix will show assymetry if the sub-population sizes are different.


```{r message=FALSE, warning=FALSE}
contact_matrix(survey = polymod, countries = "Germany",age.limits = c(0,60), symmetric = T, per.capita = TRUE)
```





